---
title: "CASPEN Tutorial: Celltype-Aware Stacked Pathway-Guided Ensemble Prediction"
author: "Shrabanti Chowdhury, Anna Xing, Pei Wang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CASPEN Tutorial: Celltype-Aware Stacked Pathway-Guided Ensemble Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

CASPEN (celltype aware stacked pathway guided ensemble) is designed for
pathway-guided prediction when the biological signal is organized by pathways,
cell types, or both. It supports three outcome families:

- binary outcomes, evaluated with AUC and sensitivity at high specificity
- survival outcomes, evaluated with time-dependent AUC, C-index, and SP95
- categorical outcomes, evaluated with multiclass AUC and SP95

The central idea is simple: build models within pathway-celltype feature sets,
select predictive pathway-celltypes, optionally refine genes within those
pathways, and then evaluate training or independent test performance.

```{r workflow-figure, echo = FALSE, out.width = "100%"}
knitr::include_graphics("../man/figures/caspen-workflow.svg")
```

## Function Map

| Function | Purpose | Typical input | Typical output |
|---|---|---|---|
| `pathway_select()` | Select predictive pathways or pathway-celltypes | outcome, omics data, pathway list, models | AUC/C-index/SP95 matrices, selected labels, priors |
| `gene_select()` | Select genes inside selected pathways | selected pathways, omics data, outcome, native-importance models | selected gene lists and ranking table |
| `Train_perform_CV()` | Estimate training performance by cross-validation | selected features and models | cross-validated model metrics |
| `Test_perform()` | Evaluate an independent test set | train/test data, selected features, models | test-set model metrics |

## Load Example Data

The package includes a compact example object. It contains a binary outcome,
an omics matrix, and a list of pathway gene sets.

```{r load-data}
library(CASPEN)
data(caspen_example)

names(caspen_example)
dim(caspen_example$data_to_pass)
table(caspen_example$outcome)
length(caspen_example$pathways)
```

For a quick tutorial, we use a small subset.

```{r subset-data}
set.seed(1)
x <- caspen_example$data_to_pass
y <- caspen_example$outcome
pathways <- caspen_example$pathways[1:3]
```

## Model Backends And Versions

CASPEN uses short model abbreviations in all three outcome families. To record
the modeling packages and installed versions used in an analysis, run:

```{r model-info}
caspen_model_info()
```

Common abbreviations are `XG` for xgboost, `RF` for random forest, `EN` for
elastic net, `GB` for gradient boosting, `KNN` for k-nearest neighbors, `SVM`
for support vector machines, `NB` for naive Bayes, `DCT` for decision trees,
`NN` for neural networks, `ADB` for AdaBoost, and `MB` for model-based
boosting.

## Step 1: Select Pathways

`pathway_select()` trains models for each pathway and returns performance
matrices. For binary outcomes, the primary metric is AUC. For survival outcomes,
the primary metric is C-index when available. SP95 is sensitivity at 95%
specificity.

```{r pathway-select-example, eval = FALSE}
path_res <- pathway_select(
  outcome.type = "binary",
  outcome = y,
  data = x,
  features = pathways,
  models.indiv = c("RF", "EN", "DCT"),
  iter = 5,
  num.folds = 3,
  ensemble = TRUE,
  models.ens = c("RF", "EN"),
  num.sel.path = 2,
  fs.AUC.cut = 0.60,
  fs.sp.cut = 0.05
)

path_res$auc
path_res$SP.95
path_res$selection.table
path_res$features
```

To parallelize repeated iterations within each pathway, turn on the future
backend. By default, CASPEN uses `future::availableCores()` and caps the worker
count at `iter`, so users do not need to hard-code a worker number.

```{r pathway-select-parallel-example, eval = FALSE}
path_res <- pathway_select(
  outcome.type = "binary",
  outcome = y,
  data = x,
  features = pathways,
  models.indiv = c("RF", "EN", "DCT"),
  iter = 100,
  num.folds = 3,
  parallel.iter = TRUE
)
```

### Adding Literature or LLM Priors

CASPEN can use user-supplied literature signatures:

- additional literature-derived gene sets through `literature.features`
- pathway prior weights through `pathway.prior`

These priors can be pathway-specific or pathway-celltype-specific.

```{r prior-example, eval = FALSE}
literature_sets <- list(
  Literature_Immune_Response = c("STAT1", "IRF1", "CXCL10", "B2M")
)

pathway_prior <- c(
  "akt" = 0.60,
  "anti_apoptosis" = 0.50,
  "Literature_Immune_Response" = 0.95
)

path_res <- pathway_select(
  outcome.type = "binary",
  outcome = y,
  data = x,
  features = pathways,
  literature.features = literature_sets,
  pathway.prior = pathway_prior,
  models.indiv = c("RF", "EN"),
  iter = 3,
  num.folds = 3
)

path_res$data.metric
path_res$SP.95
path_res$Prior.weight
```

If user-supplied signatures are not available, CASPEN can query an LLM for
disease/outcome-specific signatures and prior weights. By default this uses the
OpenAI API through `OPENAI_API_KEY` and can retrieve PubMed abstracts as
grounding context.

CASPEN supports three literature-prior routes:

1. Use `prior.method = "pubmed_count"` in `llm_literature_signatures()` or
   `llm.prior.method = "pubmed_count"` in `pathway_select()` to avoid an LLM
   call and score pathways by normalized PubMed hit counts. Use
   `prior.method = "pubtator"` or `llm.prior.method = "pubtator"` to score
   PubTator3 gene-annotation evidence.
2. Use the OpenAI API by setting `OPENAI_API_KEY` or passing `llm.api.key`.
3. Use a local/custom LLM by passing `llm.provider = "custom"` and an `llm.fn`
   function. Ollama is a convenient local option.

OpenAI setup:

```{r openai-setup-example, eval = FALSE}
Sys.setenv(OPENAI_API_KEY = "sk-proj-your_real_key_here")

llm_res <- llm_literature_signatures(
  disease = "high-grade serous ovarian cancer",
  outcome.context = "platinum resistance",
  features = pathways,
  mode = "prior",
  model = "gpt-4o-mini",
  pubmed = FALSE,
  temperature = NULL
)
```

PubMed-count prior without an LLM:

```{r pubmed-count-example, eval = FALSE}
pubmed_res <- llm_literature_signatures(
  disease = "high-grade serous ovarian cancer",
  outcome.context = "platinum resistance",
  features = pathways,
  mode = "prior",
  prior.method = "pubmed_count",
  email = "your.email@example.com"
)

pubmed_res$evidence[, c("pathway", "pubmed.count", "pubmed.score", "prior")]
```

PubTator3 prior without an LLM:

```{r pubtator-example, eval = FALSE}
pubtator_res <- llm_literature_signatures(
  disease = "high-grade serous ovarian cancer",
  outcome.context = "platinum resistance",
  features = pathways,
  mode = "prior",
  prior.method = "pubtator",
  pubmed.max = 20,
  email = "your.email@example.com"
)

pubtator_res$evidence[, c("pathway", "pubtator.count",
                          "pubtator.score", "prior")]
```

Hybrid PubMed + PubTator + LLM prior:

```{r hybrid-evidence-example, eval = FALSE}
hybrid_res <- llm_literature_signatures(
  disease = "high-grade serous ovarian cancer",
  outcome.context = "platinum resistance",
  features = pathways,
  mode = "prior",
  prior.method = "hybrid",
  hybrid.weights = c(pubmed_count = 0.25, pubtator = 0.25, llm = 0.5),
  model = "gpt-4o-mini",
  pubmed = FALSE,
  temperature = NULL
)
```

Local Ollama setup:

1. Install Ollama from <https://ollama.com/download>.
2. In Terminal, run `ollama pull qwen2.5:7b`.
3. Keep the Ollama app or service running while R calls
   `http://localhost:11434`.

```{r ollama-example, eval = FALSE}
ollama_llm <- function(prompt) {
  req <- httr2::request("http://localhost:11434/api/generate") |>
    httr2::req_body_json(list(
      model = "qwen2.5:7b",
      prompt = prompt,
      stream = FALSE
    ))
  res <- httr2::req_perform(req)
  out <- jsonlite::fromJSON(httr2::resp_body_string(res))
  out$response
}

ollama_res <- llm_literature_signatures(
  disease = "high-grade serous ovarian cancer",
  outcome.context = "platinum resistance",
  features = pathways,
  mode = "prior",
  provider = "custom",
  llm.fn = ollama_llm,
  pubmed = FALSE
)
```

```{r llm-example, eval = FALSE}
path_res <- pathway_select(
  outcome.type = "binary",
  outcome = y,
  data = x,
  features = pathways,
  models.indiv = c("RF", "EN", "DCT"),
  iter = 50,
  num.folds = 3,
  llm.mode = "both",
  disease = "high-grade serous ovarian cancer",
  outcome.context = "platinum resistance"
)

path_res$data.metric
path_res$SP.95
path_res$Prior.weight
path_res$llm$evidence
```

`llm.mode = "curate"` adds new literature signatures without prior scoring,
`llm.mode = "prior"` scores existing user pathways without adding new
signatures, and `llm.mode = "both"` does both. CASPEN keeps the model metric,
SP95, and prior weight as separate outputs so users can choose cutoffs directly.

The LLM output can also be generated and inspected before model fitting:

```{r llm-helper-example, eval = FALSE}
llm_res <- llm_literature_signatures(
  disease = "high-grade serous ovarian cancer",
  outcome.context = "platinum resistance",
  features = pathways,
  mode = "both"
)

llm_res$literature.features
llm_res$pathway.prior
llm_res$evidence
```

## Step 2: Celltype-Aware Pathway Mapping

If the columns of the omics matrix encode both gene and cell type, CASPEN can
expand bulk pathways into pathway-celltype features.

Supported column styles include:

- `GENE_Celltype`
- `Celltype_GENE`
- `GENE.Celltype`
- custom separators through `celltype.sep`

If the pathway name already contains a cell type or alias, it is mapped only to
that cell type. Otherwise, it is expanded to all supplied cell types.

```{r celltype-example, eval = FALSE}
celltypes <- c("CAF", "CD4_Tcells", "CD8_Tcells", "Macrophages")

celltype_aliases <- list(
  CD4_Tcells = c("immune", "T cells", "CD4"),
  CD8_Tcells = c("immune", "T cells", "CD8"),
  Macrophages = c("immune", "macrophage")
)

pathways_ct <- list(
  hypoxia = c("ARNT", "BCL2", "CDKN1A"),
  niche_immune = c("STAT1", "B2M", "IRF1")
)

path_res <- pathway_select(
  outcome.type = "binary",
  outcome = y,
  data = celltype_expression,
  features = pathways_ct,
  celltypes = celltypes,
  celltype.aliases = celltype_aliases,
  models.indiv = c("RF", "EN"),
  iter = 5,
  num.folds = 3
)
```

In this example, `hypoxia` is expanded to every cell type, while
`niche_immune` maps only to `CD4_Tcells`, `CD8_Tcells`, and `Macrophages`
because those cell types have `"immune"` as an alias.

## Step 3: Select Genes Within Selected Pathways

`gene_select()` is a separate optional step. It uses native model importance
from models such as `XG`, `RF`, `GB`, `EN`, `MB`, and `DCT`.

Because raw importance has different meanings in different models, CASPEN does
not average raw values. Instead it:

1. converts native importance to within-run percentiles,
2. computes the median percentile for each gene,
3. computes stability as the proportion of runs where the gene reaches the
   chosen percentile threshold,
4. filters genes by the user-specified stability cutoff,
5. returns the top genes ordered by median percentile.

The user controls the cutoffs:

- `min.pathway.size`: only pathways larger than this are reduced
- `gene.stability.quantile`: percentile threshold for stability, default 0.75
- `gene.min.stability`: minimum stability required, default 0.25
- `max.genes`: maximum number of genes to keep after filtering

```{r gene-select-example, eval = FALSE}
gene_res <- gene_select(
  outcome.type = "binary",
  outcome = y,
  data = x,
  features = path_res$features,
  models = c("RF", "EN", "DCT"),
  iter = 10,
  num.folds = 3,
  min.pathway.size = 30,
  gene.stability.quantile = 0.75,
  gene.min.stability = 0.25,
  max.genes = 30
)

gene_res$selected.features
head(gene_res$ranking)
```

If no gene in a pathway passes `gene.min.stability`, the selected gene list for
that pathway is empty. This makes the cutoff behavior explicit and user-driven.

## Step 4: Training Performance

After selecting pathway-celltypes or genes, use `Train_perform_CV()` to estimate
cross-validated training performance.

```{r train-performance-example, eval = FALSE}
train_res <- Train_perform_CV(
  outcome.type = "binary",
  outcome = y,
  train_data = x,
  features = gene_res$selected.features,
  models.indiv = c("RF", "EN", "DCT"),
  iter = 10,
  ensemble = TRUE,
  models.ens = c("RF", "EN")
)

train_res
```

## Step 5: Independent Test Performance

For an external validation cohort, use `Test_perform()`. The same selected
features should be supplied for training and testing.

```{r test-performance-example, eval = FALSE}
test_res <- Test_perform(
  outcome.type = "binary",
  outcome = y_train,
  train_data = x_train,
  test_data = x_test,
  test_outcome = y_test,
  features = gene_res$selected.features,
  models.indiv = c("RF", "EN", "DCT"),
  ensemble = TRUE,
  models.ens = c("RF", "EN")
)

test_res
```

## Survival Outcomes

For survival prediction, provide both event status and survival time.
`pathway_select()` uses C-index as the primary pathway-selection metric when
available, while time-dependent AUC and SP95 are also returned.

```{r survival-example, eval = FALSE}
surv_res <- pathway_select(
  outcome.type = "survival",
  outcome = status,
  survdays = days,
  time_point = 365,
  data = x,
  features = pathways,
  models.indiv = c("RF", "EN", "GB"),
  iter = 5,
  num.folds = 3,
  fs.AUC.cut = 0.60,
  fs.sp.cut = 0.05
)

surv_res$C.index
surv_res$auc
surv_res$SP.95
```

## Categorical Outcomes

For categorical outcomes, CASPEN supports multiclass prediction using models
such as random forest, elastic net, gradient boosting, model-based boosting, and
one-vs-rest learners where needed.

```{r categorical-example, eval = FALSE}
cat_res <- pathway_select(
  outcome.type = "categorical",
  outcome = subtype,
  data = x,
  features = pathways,
  models.indiv = c("RF", "EN", "MB", "DCT"),
  iter = 5,
  num.folds = 3
)

cat_res$auc
cat_res$SP.95
```

## Practical Recommendations

- Start with pathway-level selection before gene-level selection.
- Use celltype aliases to make broad terms such as `"immune"` explicit.
- Use literature or LLM priors as external inputs, not as hidden training-time
  dependencies.
- For large pathways, run `gene_select()` only after pathway filtering.
- Keep the gene-selection cutoff user-driven: stability determines eligibility,
  and median percentile determines ranking.
- For final reporting, inspect both `selected.features` and the full `ranking`
  table.

## Minimal End-to-End Skeleton

```{r skeleton, eval = FALSE}
path_res <- pathway_select(
  outcome.type = "binary",
  outcome = y,
  data = x,
  features = pathways,
  models.indiv = c("RF", "EN", "DCT"),
  iter = 10,
  num.folds = 3,
  num.sel.path = 10
)

gene_res <- gene_select(
  outcome.type = "binary",
  outcome = y,
  data = x,
  features = path_res$features,
  models = c("RF", "EN", "DCT"),
  iter = 10,
  num.folds = 3,
  min.pathway.size = 30,
  gene.stability.quantile = 0.75,
  gene.min.stability = 0.25,
  max.genes = 30
)

train_res <- Train_perform_CV(
  outcome.type = "binary",
  outcome = y,
  train_data = x,
  features = gene_res$selected.features,
  models.indiv = c("RF", "EN", "DCT"),
  iter = 10,
  ensemble = TRUE,
  models.ens = c("RF", "EN")
)
```
